A supertree pipeline for summarizing phylogenetic and taxonomic information for millions of species

### 2016-10-12 doc/paper1-submission/paper4.tex
final MTH edits

We compare our supertree construction method to a previously published
supertree construction method by assessing their performance on input
trees used to construct the Open Tree of Life version 4, and find that
our method increases the number of displayed input splits from 35,518
to 39,639 and decreases the number of conflicting input splits from
2760 to 1357.  The new supertree method also improves on the previous
supertree construction method in that it produces no unsupported
branches and avoids unnecessary polytomies.

@@

We can consider whether or not an node in an input tree is displayed
by $\summaryTree$. For any such node $j$ there is a set of taxa
that are mapped to the tips that descend from the node. This set of
taxa is the ``cluster'' of taxa corresponding to node $j$; it can
be denoted $\mathcal{L}(i,j)$. It will also be referred to as the
``include set'' of the node. The ``exclude set'' of $j$ in $T_{i}$
is the set of taxa in $\mathcal{L}(i)$ but not in $\mathcal{L}(i,j)$.
If the cluster of taxa for any supertree node in $\summaryTree(i)$
is identical to $\mathcal{L}(i,j)$, then we say that $\summaryTree$
displays node $j$ of $T_{i}$. We say that the summary tree \emph{displays}
edge $j$ if the summary tree displays the child node of edge $j$.
Operationally, we can find the most recent common ancestor (MRCA)
node of $\mathcal{L}(i,j)$ in $\summaryTree$; the summary tree displays
$j$ if and only if that MRCA node is not an ancestor of any member
of the exclude set of $j$. We say that edge $k$ of the summary tree
is \emph{supported by} node $j$ in $T_{i}$ if the summary tree displays
node $j$, but if we contracted edge $k$ then the modified summary
tree would no longer display node $j$.

@@

In order to produce a comprehensive supertree, we also require a rooted
taxonomy tree in addition to the ranked list of rooted input trees.
Unlike other input trees, the taxonomy tree is required to contain
all taxa, and thus has the maximal leaf set. We make the taxonomy
tree the lowest ranked tree. In our current formulation, the taxonomy
tree is also unique in that the taxonomy is the only source of taxonomic
names. Each node in the taxonomy tree corresponds to a named group.
Taxonomic groups may have the same name, but each node in the taxonomy
tree is identified by a unique number (its OTT Id). Taxonomic groups
are identified in the summary supertree by finding a branch (or ``node'')
that has exactly the same include|exclude relationship. The taxonomy
supertree can meaningfully possess degree-two nodes. Although these
nodes can be removed without affecting the relationships of the leaves,
they do represent nested taxonomic groups that contain exactly one
subgroup. The taxonomy is also used to determine which tips are terminal
taxa.

@@

The supertree should be as resolved as possible - in other words,
it should have no unnecessary polytomies. Thus, for each input edge
that is \emph{not} included, we can point to a reason for non-inclusion
by showing that the input edge conflicts with some edge of the summary
tree. Note, that the requirement to not display unsupported groups
leads to some ``necessary'' polytomies. For example any resolution
of the polytomy shown in figure (\ref{fig:pairwisecompat}e) would
continue to display the same two input groups, but the additional
grouping would be unsupported, because the unresolved tree already
displays the two input groups. Thus, the additional group would be
unsupported and the unresolved tree would be preferred by our criterion.
However, collapsing either internal edge of the tree shown in figure
(\ref{fig:pairwisecompat}d) would result in a tree which displays
only one input grouping. This tree would contain an unnecessary polytomy,
because the polytomy would permit refinement to the depicted tree
which displays more input groupings.

@@

OTT is a hierarchy of taxonomic names that implies a phylogenetic
taxonomy. An OTT ID has a position in the hierarchy, a taxonomic name,
and set of references to the same name in different taxonomies. In
addition, the ID may also be associated with a set of flags that can
indicate that the taxon may be questionable. These flags can either
encode information taken from an input taxonomy (for example, taxa
the NCBI refers to as ``unplaced'' are assigned an ``unplaced''
flag) or can arise because of some form of conflict during taxonomy
construction (for example, if two taxonomies disagree on the name
for a taxon, then the taxon can be merged and the name will be retained
without any descendants; this name will have an OTT Id, but will be
flagged as ``barren''). Propinquity prunes the OTT down to a more
reliable taxonomy by pruning off parts of the tree that are flagged
with suspicious flags. The set of flags that lead to a subtree of
the taxonomy being pruned is under the control of the user (the set
of flags used by the Open Tree of Life project can be found in the
\texttt{config.opentree.synth} file in the propinquity repository).
For the purpose of the rest of the pipeline, an OTT Id that has been
pruned from the taxonomy will be treated in the same way as invalid
OTT Id. The output of this step is stored in propinquity's \texttt{cleaned\_ott}
subdirectory; this operation only needs to be performed when the OTT
or the pruning flags change.

@@

Many input trees have tips that are not terminal taxa, but higher-order
taxonomic groups. It is not clear how to interpret a tip in a phylogenetic
estimate that is labeled with the name of a higher taxon. Several
scenarios can lead to these cases: the data for the tip could have
been created by merging a chimeric set of character scores from constituent
taxa; the species sampled may not have been identified to the lowest
taxonomic rank; or the researcher may simply have used a higher taxonomic
name because he/she assumed that the taxon is monophyletic and the
higher level name would be more recognizable. Rather than allowing
the ambiguity about interpretation of the higher-taxon mapped tips
to propagate throughout the entire pipeline, we transform the input
trees by replacing higher taxa at tips with a set of terminal-taxon
exemplars for each taxon. One approach would be to simply determine
all descendant terminal taxa and attach them as children of the problematic
tip. However, this would create a clade rather than a tip; subsequent
steps in the supertree would interpret the clade as a claim of monophyly
for the taxon. The input tree may not have tested monophyly of the
clade, so this interpretation is unwarranted. We avoid it by attaching
exemplar taxa as child nodes of the higher taxonomic tip but then
collapsing the edge between the node that connected the higher taxon
to its parent. Thus, if $A$ is a non-terminal taxon containing terminal
descendants $a_{1}$ and $a_{2}$ and $B$ is a non-terminal taxon
containing terminal descendants $b_{1}$ and $b_{2}$ we would replace
the subtree $((A,B),c)$ with $((a_{1},a_{2},b_{1},b_{2}),c)$ instead
of the subtree $(((a_{1},a_{2})A,(b_{1},b_{2})B),c)$.

@@

After the preprocessing steps, the inputs have been converted to a
set of rooted phylogenetic estimates in which each leaf is mapped
to a terminal taxon in the exemplified taxonomic tree. The goal of
the remainder of the pipeline is to construct a tree that maximizes
the sum of displayed groups' weighted scores criterion. This is accomplished
in four steps: (1) dividing the full problem into subproblems based
on uncontested taxa, (2) constructing a summary solution for each
subproblem by greedily creating a maximally-sized list of groupings
that can all be displayed simultaneously; (3) grafting the subproblem
solutions into a single supertree; and (4) grafting (or ``unpruning'')
the taxonomy-only taxa onto the solution to produce a complete summary
tree.

@@

Note that decomposition into uncontested groups does not necessarily
allow us to find the tree that maximizes the MSDGWS score. For example,
see figure \ref{fig:decompose-worsens}; that example is a variant
of the situation shown in figure \ref{fig:pairwisecompat}. In this
case the groupings from each of the phylogenetic estimates, shown
in panels \ref{fig:decompose-worsens}a and (\ref{fig:decompose-worsens}b,
could be displayed. That solution is shown in panel \ref{fig:decompose-worsens}d,
it displays two of the three input splits, but is optimal because
no solution displays all three input groupings and the depicted solution
displays the two highest ranked groupings. However, neither of the
trees shown in panels \ref{fig:decompose-worsens}a or \ref{fig:decompose-worsens}b
contest the taxon $B$ shown in the taxonomy panel \ref{fig:decompose-worsens}c.
Thus, when using our decomposition, the branches leading to taxa $B1$
and $B2$ in the input phylogenetic trees would be sliced during the
decomposition, and relabeled to refer to taxon $B$. This taxonomically-informed
interpretation of the inputs views the two phylogenetic inputs as
in conflict; so the solution returned by propinquity would defer to
the higher ranked tree. The tree shown in panel \ref{fig:decompose-worsens}e
would be returned. This example arises from the fact that the trees
in \ref{fig:decompose-worsens}a and \ref{fig:decompose-worsens}b
jointly contest taxon $B$, but neither contests taxon $B$ when the
trees are considered in isolation.

@@

Figure \ref{fig:Two-approaches-to-unpruning} illustrates the two
approaches to unpruning. Taxa G, M, and R (Fig \ref{fig:Two-approaches-to-unpruning}a)
are broken because they conflict with the grafted solution (Fig \ref{fig:Two-approaches-to-unpruning}b);
among these, only taxon R has children that were unsampled in the
grafted solution. Ignoring all broken taxa when unpruning would cause
the unsampled children (R4, R5, and R6) to attached directly at taxon
N (as in the tree shown in Fig \ref{fig:Two-approaches-to-unpruning}c),
because that is the least inclusive unbroken ancestor of R. The tree
illustrated in Fig \ref{fig:Two-approaches-to-unpruning}d shows the
tree that would be produced by propinquity; the children of the broken
taxon R and instead attached at the MRCA of sampled children (R1,
R2, and R3). Their attachment point does not correspond to a taxon
in the taxonomic tree.

@@

In order to annotate each node in the summary supertree, it is first
necessary that each node have a unique identifier. Nodes whose include
group correspond exactly to the include group of a node in the taxonomy
are given the same identifier as the corresponding taxonomy node.
These identifiers are of the form $ottX$ where $X$ is an integer
OTT id. We generate a label of the form $mrcaottX_{1}ottX_{2}$ for
an non-taxonomic node $n$ where $X_{1}$ and $X_{2}$ are the OTT
ids for two leaves, $n$ is the MRCA of these leaves, and $X_{1}$
is the numerically smallest OTT ID that is a descendant of $n$, and
$X_{2}$ is the next the smallest ID that can be chosen to designate
$n$ as the MRCA. Because new taxa added to OTT will be given higher
OTT Ids, the use of the lowest numbered OTT Ids as designators increases
the chance that a node label can be encountered in a subsequent version
of the tree (though the taxonomic content may change). The deterministic
choice of designators also makes the labeling insensitive to branch
rotation of the grafted solution tree.

@@

\caption{\label{fig:displayed-annotations}The relationship of edges in summary
tree $\protect\summaryTree$ (b) to edges in the input tree $T_{1}$
(a). Only edges of $\protect\summaryTree$ that are present in the
induced tree $\protect\summaryTree(1)$ are represented by JSON annotations
(d). Taxon names are here suppressed in favor of OTT ids, and edges
are referenced via their tipward nodes. Edges in $\protect\summaryTree(1)$
that correspond to terminal edges of $T_{1}$ are orange; edges of
$\protect\summaryTree(1)$ that are supported by edges of $T_{1}$
are blue; where multiple edges of $\protect\summaryTree(1)$ correspond
to the same edge of $T_{1}$ they are green. There is no conflict
in this example. Also, if this were output from propinquity, then
each internal node of $\protect\summaryTree$ would be supported by
other inputs trees that are not shown here. }
\end{figure}

@@

The flag-cleaned version of OTT contained 2,424,255 leaves.  The
\otl{} supertree was constructed from 482 phylogenetic inputs,
containing a total of 45385 leaves, of which 41029 were unique.
After flag-cleaning and exemplification by propinquity, these trees contained 40323
unique tips. We used the same cleaning flags to trim the taxonomy and
input trees when constructing \otlprop{}, so \otlprop{} contains the
same number of leaves.

@@

% conflicts in v4: 2760
% conflicts in v4propr: 1357

@@

\caption{Representation of taxonomy splits in the \otl{} tree and the \otlprop{} tree.}
\resizebox{\textwidth}{!}{%
\begin{tabular}{lcccccc}
 & \supportedby{} & \partialpathof{} & \terminal{} & \conflictswith{} & \resolvedby{} & \resolves{}\tabularnewline
\hline 
\otl{} & 125,384 & 0 & 2,424,255 & 1998 & 4 & 3676\tabularnewline
\hline 
\otl{} only & 296 & 0 & 0 & 19 & 4 & 17\tabularnewline
\hline 
\otlprop{} & 125,107 & 0 & 2,424,255 & 2,279 & 0 & 3,883\tabularnewline
\hline 
\otlprop{} only & 19 & 0 & 0 & 300 & 0 & 224\tabularnewline
\hline 
\label{tab:representtaxonomy}
\end{tabular}}

@@

We performed an annotation of both the \otl{} tree and the \otlprop{}
tree as described in section \ref{sec:Annotation} to assess the
ability of our new supertree method to represent splits from input
phylogenies.  Table \ref{tab:representinputs} classifies the input
phylogeny splits according to how they relate to a summmary tree, so
that each input edge falls in one of \supportedby{}, \partialpathof{},
\terminal{}, \conflictswith{}, or \resolvedby{}. For example, the
numbers in the \conflictswith{} column indicate the number of input
splits $j$ with at least one summary edge $x$ such that the relation
``$x$ \conflictswith{} $j$'' holds.  The total number of non-terminal
input phylogeny splits considered thus 40996.

The number of displayed input splits (\supportedby{}
+ \partialpathof{}) increased from 35,518 (for \otl{}) to 39,639 (for \otlprop{});
an 11.2\% increase. When examining which splits are displayed, we find
that the \otlprop{} tree displays 4920 input splits that
are not displayed by the \otl{} tree, whereas the \otl{} tree displays
only 799 input splits that are not displayed by the \otlprop{} tree.  The number of input
splits that conflict with the summary (conflicts\_with) dropped from
2760 to 1357, a decrease of 1357, or 49.2\%.  In accordance with the
goal of not containing unnecessary polytomies, the number of input
splits that do not conflict with the summary tree, but are not
incorporated (resolved\_by) dropped from 2718 to 0.  We also find that
the number of polytomies in input phylogenies that are resolved by the
summary tree increases from 473 for \otl{} to 515 for \otlprop{}.

@@

Here we have described the motivation and methodology used by our new
supertree method that is currently used by the Open Tree of Life
project to build its summary tree.  Our new method represented 11\%
more input phylogeny splits with 49\% less conflict compared to the
Open Tree of Life version 4 summary tree when applied to the same inputs.
Unless the previous method \citep{HinchliffEtAl2015}, our new method
is guaranteed to incorporate input splits that do not conflict with
the summary tree. The method is implemented in the Open Source
software package propinquity.  A modified version of the treemachine
software which built the summary tree described in the
\citet{HinchliffEtAl2015} paper is used by the project to serve the
tree produced by propinquity via Open Tree of Life APIs.

@@

Thanks to Karen Cranston, Jonathan Rees, Jim Allman, Cody Hinchliff,
Stephen Smith, and Joseph Brown for discussions and feedback. Thanks
to NSF grant 1208393 (DEB), part of the collaborative Open Tree of
Life awards, the University of Kansas, and the Heidelberg Institute
for Theoretical Studies for funding.

>>>>

We compare our supertree construction method to a previously published
supertree construction method by assessing their performance on input
trees used to construct the Open Tree of Life version 4, and find that
our method increases the number of displayed input splits from 35,518
to 39,639 and decreases the number of conflicting input splits from
2,760 to 1,357.  The new supertree method also improves on the previous
supertree construction method in that it produces no unsupported
branches and avoids unnecessary polytomies.

@@

We can consider whether or not a node in an input tree is displayed
by $\summaryTree$. For any such node $j$ there is a set of taxa
that are mapped to the tips that descend from the node. This set of
taxa is the ``cluster'' of taxa corresponding to node $j$; it can
be denoted $\mathcal{L}(i,j)$. It will also be referred to as the
``include set'' of the node. The ``exclude set'' of $j$ in $T_{i}$
is the set of taxa in $\mathcal{L}(i)$ but not in $\mathcal{L}(i,j)$.
If the cluster of taxa for any supertree node in $\summaryTree(i)$
is identical to $\mathcal{L}(i,j)$, then we say that $\summaryTree$
displays node $j$ of $T_{i}$. We say that the summary tree \emph{displays}
edge $j$ if the summary tree displays the child node of edge $j$.
Operationally, we can find the most recent common ancestor (MRCA)
node of $\mathcal{L}(i,j)$ in $\summaryTree$; the summary tree displays
$j$ if and only if that MRCA node is not an ancestor of any member
of the exclude set of $j$. We say that node $k$ of the summary tree
is \emph{supported by} node $j$ in $T_{i}$ if the summary tree displays
node $j$, but if we contracted edge that separates $k$ from its parent then the modified summary
tree would no longer display node $j$.

@@

In order to produce a comprehensive supertree, we also require a rooted
taxonomy tree in addition to the ranked list of rooted input trees.
Unlike other input trees, the taxonomy tree is required to contain
all taxa, and thus has the maximal leaf set. We make the taxonomy
tree the lowest ranked tree. In our current formulation, the taxonomy
tree is also unique in that the taxonomy is the only source of taxonomic
names. Each node in the taxonomy tree corresponds to a named group.
Taxonomic groups may have the same name, but each node in the taxonomy
tree is identified by a unique number (its OTT ID). Taxonomic groups
are identified in the summary supertree by finding a branch (or ``node'')
that has exactly the same include|exclude relationship. The taxonomy
supertree can meaningfully possess degree-two nodes. Although these
nodes can be removed without affecting the relationships of the leaves,
they do represent nested taxonomic groups that contain exactly one
subgroup. The taxonomy is also used to determine which tips are terminal
taxa.

@@

The supertree should be as resolved as possible - in other words,
it should have no unnecessary polytomies. Thus, for each input edge
that is \emph{not} included, we can point to a reason for non-inclusion
by showing that the input edge conflicts with some edge of the summary
tree. Note, that the requirement to not display unsupported groups
leads to some ``necessary'' polytomies. For example any resolution
of the polytomy shown in figure (\ref{fig:pairwisecompat}e) would
continue to display the same two input groups. However, the additional
grouping would be unsupported, because the unresolved tree already
displays both input groups. Thus, the 
unresolved tree would be preferred by our criterion.
However, collapsing either internal edge of the tree shown in figure
(\ref{fig:pairwisecompat}d) would result in a tree which displays
only one input grouping. This tree would contain an unnecessary polytomy,
because the polytomy would permit refinement to the depicted tree
which displays more input groupings.

@@

OTT is a hierarchy of taxonomic names that implies a phylogenetic
taxonomy. An OTT ID has a position in the hierarchy, a taxonomic name,
and set of references to the same name in different taxonomies. In
addition, the ID may also be associated with a set of flags that can
indicate that the taxon may be questionable. These flags can either
encode information taken from an input taxonomy (for example, taxa
the NCBI refers to as ``unplaced'' are assigned an ``unplaced''
flag) or can arise because of some form of conflict during taxonomy
construction (for example, if two taxonomies disagree on the name
for a taxon, then the taxon can be merged and the name will be retained
without any descendants; this name will have an OTT ID, but will be
flagged as ``barren''). Propinquity prunes the OTT down to a more
reliable taxonomy by pruning off parts of the tree that are flagged
with suspicious flags. The set of flags that lead to a subtree of
the taxonomy being pruned is under the control of the user (the set
of flags used by the Open Tree of Life project can be found in the
\texttt{config.opentree.synth} file in the propinquity repository).
For the purpose of the rest of the pipeline, an OTT ID that has been
pruned from the taxonomy will be treated in the same way as invalid
OTT ID. The output of this step is stored in propinquity's \texttt{cleaned\_ott}
subdirectory; this operation only needs to be performed when the OTT
or the pruning flags change.

@@

Many input trees have tips that are not terminal taxa, but higher-order
taxonomic groups. It is not clear how to interpret a tip in a phylogenetic
estimate that is labeled with the name of a higher taxon. Several
scenarios can lead to these cases: the data for the tip could have
been created by merging a chimeric set of character scores from constituent
taxa; the species sampled may not have been identified to the lowest
taxonomic rank; or the researcher may simply have used a higher taxonomic
name because he/she assumed that the taxon is monophyletic and the
higher level name would be more recognizable. Rather than allowing
the ambiguity about interpretation of the higher-taxon mapped tips
to propagate throughout the entire pipeline, we transform the input
trees by replacing higher taxa at tips with a set of terminal-taxon
exemplars for each taxon. One approach would be to simply determine
all descendant terminal taxa and attach them as children of the problematic
tip. However, this would create a clade rather than a tip; subsequent
steps in the supertree would interpret the clade as a claim of monophyly
for the taxon. The input tree may not have tested monophyly of the
clade, so this interpretation is unwarranted. We avoid it by attaching
exemplar taxa as child nodes of the higher taxonomic tip but then
collapsing the edge between the former tip node and
its parent. Thus, if $A$ is a non-terminal taxon containing terminal
descendants $a_{1}$ and $a_{2}$ and $B$ is a non-terminal taxon
containing terminal descendants $b_{1}$ and $b_{2}$ we would replace
the subtree $((A,B),c)$ with $((a_{1},a_{2},b_{1},b_{2}),c)$ instead
of the subtree $(((a_{1},a_{2})A,(b_{1},b_{2})B),c)$.

@@

After the preprocessing steps, the inputs have been converted to a
set of rooted phylogenetic estimates in which each leaf is mapped
to a terminal taxon in the exemplified taxonomic tree. The goal of
the remainder of the pipeline is to construct a tree that maximizes
the sum of displayed groups' weighted scores (MSDGWS) criterion. This is accomplished
in four steps: (1) dividing the full problem into subproblems based
on uncontested taxa, (2) constructing a summary solution for each
subproblem by greedily creating a maximally-sized list of groupings
that can all be displayed simultaneously; (3) grafting the subproblem
solutions into a single supertree; and (4) grafting (or ``unpruning'')
the taxonomy-only taxa onto the solution to produce a complete summary
tree.

@@

Note that decomposition into uncontested groups does not necessarily
allow us to find the tree that maximizes the MSDGWS score. For example,
see figure \ref{fig:decompose-worsens}; that example is a variant
of the situation shown in figure \ref{fig:pairwisecompat}. In this
case the groupings from each of the phylogenetic estimates, shown
in panels \ref{fig:decompose-worsens}a and \ref{fig:decompose-worsens}b,
could be displayed. That solution is shown in panel \ref{fig:decompose-worsens}d,
it displays two of the three input splits, but is optimal because
no solution displays all three input groupings and the depicted solution
displays the two highest ranked groupings. However, neither of the
trees shown in panels \ref{fig:decompose-worsens}a or \ref{fig:decompose-worsens}b
contest the taxon $B$ shown in the taxonomy panel \ref{fig:decompose-worsens}c.
Thus, when using our decomposition, the branches leading to taxa $B1$
and $B2$ in the input phylogenetic trees would be sliced during the
decomposition, and relabeled to refer to taxon $B$. This taxonomically-informed
interpretation of the inputs views the two phylogenetic inputs as
in conflict; so the solution returned by propinquity would defer to
the higher ranked tree. The tree shown in panel \ref{fig:decompose-worsens}e
would be returned. This example arises from the fact that the trees
in \ref{fig:decompose-worsens}a and \ref{fig:decompose-worsens}b
jointly contest taxon $B$, but neither contests taxon $B$ when the
trees are considered in isolation.

@@

Figure \ref{fig:Two-approaches-to-unpruning} illustrates the two
approaches to unpruning. Taxa G, M, and R (Fig \ref{fig:Two-approaches-to-unpruning}a)
are broken because they conflict with the grafted solution (Fig \ref{fig:Two-approaches-to-unpruning}b);
among these, only taxon R has children that were unsampled in the
grafted solution. Ignoring all broken taxa when unpruning would cause
the unsampled children (R4, R5, and R6) to attached directly at taxon
N (as in the tree shown in Fig \ref{fig:Two-approaches-to-unpruning}c),
because that is the least inclusive unbroken ancestor of R. The tree
illustrated in Fig \ref{fig:Two-approaches-to-unpruning}d shows the
tree that would be produced by propinquity; the children of the broken
taxon R and instead attached at the MRCA of sampled children (R1,
R2, and R3). Their attachment point does not correspond to any taxon
in the taxonomic tree.

@@

In order to annotate each node in the summary supertree, it is first
necessary that each node have a unique identifier. Nodes whose include
group correspond exactly to the include group of a node in the taxonomy
are given the same identifier as the corresponding taxonomy node.
These identifiers are of the form $ottX$ where $X$ is an integer
OTT ID. We generate a label of the form $mrcaottX_{1}ottX_{2}$ for
an non-taxonomic node $n$ where $X_{1}$ and $X_{2}$ are the OTT
IDs for two leaves, $n$ is the MRCA of these leaves, and $X_{1}$
is the numerically smallest OTT ID that is a descendant of $n$, and
$X_{2}$ is the next the smallest ID that can be chosen to designate
$n$ as the MRCA. Because new taxa added to OTT will be given higher
OTT IDs, the use of the lowest numbered OTT IDs as designators increases
the chance that a node label can be encountered in a subsequent version
of the tree (though the taxonomic content may change). The deterministic
choice of designators also makes the labeling insensitive to branch
rotation of the grafted solution tree.

@@

\caption{\label{fig:displayed-annotations}The relationship of edges in summary
tree $\protect\summaryTree$ (b) to edges in the input tree $T_{1}$
(a). Only edges of $\protect\summaryTree$ that are present in the
induced tree $\protect\summaryTree(1)$ are represented by JSON annotations
(d). Taxon names are here suppressed in favor of OTT IDs, and edges
are referenced via their tipward nodes. Edges in $\protect\summaryTree(1)$
that correspond to terminal edges of $T_{1}$ are orange; edges of
$\protect\summaryTree(1)$ that are supported by edges of $T_{1}$
are blue; where multiple edges of $\protect\summaryTree(1)$ correspond
to the same edge of $T_{1}$ they are green. There is no conflict
in this example. Also, if this were output from propinquity, then
each internal node of $\protect\summaryTree$ would be supported by
other inputs trees that are not shown here. }
\end{figure}

@@

The flag-cleaned version of OTT contained 2,424,255 leaves.  The
\otl{} supertree was constructed from 482 phylogenetic inputs,
containing a total of 45,385 leaves, of which 41,029 were unique.
After flag-cleaning and exemplification by propinquity, these trees contained 40,323
unique tips. We used the same cleaning flags to trim the taxonomy and
input trees when constructing \otlprop{}, so \otlprop{} contains the
same number of leaves.

@@

% conflicts in v4: 2,760
% conflicts in v4propr: 1,357

@@

\caption{Representation of taxonomy splits in the \otl{} tree and the \otlprop{} tree.}
\resizebox{\textwidth}{!}{%
\begin{tabular}{lcccccc}
 & \supportedby{} & \partialpathof{} & \terminal{} & \conflictswith{} & \resolvedby{} & \resolves{}\tabularnewline
\hline 
\otl{} & 125,384 & 0 & 2,424,255 & 1,998 & 4 & 3,676\tabularnewline
\hline 
\otl{} only & 296 & 0 & 0 & 19 & 4 & 17\tabularnewline
\hline 
\otlprop{} & 125,107 & 0 & 2,424,255 & 2,279 & 0 & 3,883\tabularnewline
\hline 
\otlprop{} only & 19 & 0 & 0 & 300 & 0 & 224\tabularnewline
\hline 
\label{tab:representtaxonomy}
\end{tabular}}

@@

We performed an annotation of both the \otl{} tree and the \otlprop{}
tree as described in section \ref{sec:Annotation} to assess the
ability of our new supertree method to represent splits from input
phylogenies.  Table \ref{tab:representinputs} classifies the input
phylogeny splits according to how they relate to a summmary tree, so
that each input edge falls in one of \supportedby{}, \partialpathof{},
\terminal{}, \conflictswith{}, or \resolvedby{}. For example, the
numbers in the \conflictswith{} column indicate the number of input
splits $j$ with at least one summary edge $x$ such that the relation
``$x$ \conflictswith{} $j$'' holds.  The total number of non-terminal
input phylogeny splits considered was 40,996.

The number of displayed input splits (\supportedby{}
+ \partialpathof{}) increased from 35,518 (for \otl{}) to 39,639 (for \otlprop{});
an 11.2\% increase. When examining which splits are displayed, we find
that the \otlprop{} tree displays 4,920 input splits that
are not displayed by the \otl{} tree, whereas the \otl{} tree displays
only 799 input splits that are not displayed by the \otlprop{} tree.  The number of input
splits that conflict with the summary (conflicts\_with) dropped from
2,760 to 1,357, a decrease of 1,357, or 49.2\%.  In accordance with the
goal of not containing unnecessary polytomies, the number of input
splits that do not conflict with the summary tree, but are not
incorporated (resolved\_by) dropped from 2,718 to 0.  We also find that
the number of polytomies in input phylogenies that are resolved by the
summary tree increases from 473 for \otl{} to 515 for \otlprop{}.

@@

Here we have described the motivation and methodology used by our new
supertree method that is currently used by the Open Tree of Life
project to build its summary tree.  Our new method represented 11\%
more input phylogeny splits with 49\% less conflict compared to the
Open Tree of Life version 4 summary tree when applied to the same inputs.
Unlike the previous method \citep{HinchliffEtAl2015}, our new method
is guaranteed to incorporate input splits that do not conflict with
the summary tree. The method is implemented in the Open Source
software package propinquity.  A modified version of the treemachine
software which built the summary tree described in the
\citet{HinchliffEtAl2015} paper is used by the project to serve the
tree produced by propinquity via Open Tree of Life APIs.

@@

Thanks to Emily Jane McTavish, Karen Cranston, Jonathan Rees, Jim Allman, Cody Hinchliff,
Stephen Smith, and Joseph Brown for discussions and feedback. Thanks
to NSF grant 1208393 (DEB), part of the collaborative Open Tree of
Life awards, the University of Kansas, and the Heidelberg Institute
for Theoretical Studies for funding.
